The guts of the dispRity package

### 2020-05-01 inst/gitbook/06_guts.Rmd
added examples for dispRity utilities matrix.dispRity
gitbook updates

## `dispRity` utilities

@@

### `dispRity` object utilities  <a name="dispRity.utilities"></a>
The first set of utilities are functions for manipulating `dispRity` objects:

#### `make.dispRity`
This function creates empty `dispRity` objects.

@@

####  `fill.dispRity`
This function initialises a `dispRity` object and generates its call properties.

@@

####  `matrix.dispRity`
This function extracts a specific matrix from a disparity object.
The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.

@@

#### `get.subsets`
This function creates a dispRity object that contains only elements from one specific subsets.

@@

#### `combine.subsets`
This function allows to merge different subsets.
If the

@@

#### `extract.dispRity`
This function extracts the calculated disparity values of a specific matrix.

@@

#### `rescale.dispRity`
This is the modified S3 method for `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object and can take optional arguments (for example the rescaling by dividing by a maximum value).

@@

#### `sort.dispRity`
This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.

@@

## The `dispRity` object content
The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:

 * `$matrix`: an object of class `matrix`, the full multidimensional space.
 * `$call`: an object of class `list` containing information on the `dispRity` object content.
 * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
 * `$disparity`: an object of class `list` containing the disparity values.

The `dispRity` object is loosely based on `C` structure objects.
In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint:  at any time, only one matrix is present in the `R` environment rather than multiple copies of it for each subset.

### `$matrix`
This is the multidimensional space, stored in the `R` environment as a `matrix` object.
It requires row names but not column names.
By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
This element of the `dispRity` object is never modified.

### `$call`
This element contains the information on the `dispRity` object content.
It is a `list` that can contain the following:

 * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`). This element generated only once via `chrono.subsets()` and `custom.subsets()`.
 * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use. This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
 * `$call$bootstrap`: this is a `list` containing three elements:
    * `[[1]]`: the number of bootstrap replicates (`numeric`)
    * `[[2]]`: the bootstrap method (`character`)
    * `[[3]]`: the rarefaction levels (`numeric` vector)
 * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`. These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).

### `$subsets`
This element contain the eventual subsets of the multidimensional space.
It is a `list` of subset names.
Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
This `elements` matrix is the raw (observed) elements in the subsets.
The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:

@@

### `$disparity`
The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:

@@

<!-- ## Modularity

### Making your own metrics

### Making your own tests

## Where is this going?
 -->
<!-- ## Running `dispRity` in parallel

The computationally intensive function `dispRity` have a `parallel` option to speed up their calculations.

This option requires the package `snow` and takes arguments that are to be passed to  -->

>>>>

## `dispRity` utilities {#utilities}

@@

### `dispRity` object utilities  <a name="dispRity.utilities"></a>

The first set of utilities are functions for manipulating `dispRity` objects:

#### `make.dispRity`

This function creates empty `dispRity` objects.

@@

####  `fill.dispRity`

This function initialises a `dispRity` object and generates its call properties.

@@

####  `matrix.dispRity`

This function extracts a specific matrix from a disparity object.
The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.

@@

#### `get.subsets`

This function creates a dispRity object that contains only elements from one specific subsets.

@@

#### `combine.subsets`

This function allows to merge different subsets.

@@

#### `extract.dispRity`

This function extracts the calculated disparity values of a specific matrix.

@@

#### `rescale.dispRity`

This is the modified S3 method for `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object and can take optional arguments (for example the rescaling by dividing by a maximum value).

@@

#### `sort.dispRity`

This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.

@@

## The `dispRity` object content

The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:

 * `$matrix`: an object of class `list` that contains at least one object of class `matrix`: the full multidimensional space.
 * `$call`: an object of class `list` containing information on the `dispRity` object content.
 * `$subsets`: an object of class `list` containing the subsets of the multidimensional space.
 * `$disparity`: an object of class `list` containing the disparity values.

The `dispRity` object is loosely based on `C` structure objects.
In fact, it is composed of one unique instance of a matrix (the multidimensional space) upon which the metric function is called via "pointers" to only a certain number of elements and/or dimensions of this matrix.
This allows for: (1) faster and easily tractable execution time: the metric functions are called through apply family function and can be parallelised; and (2) a really low memory footprint: at any time, only one matrix (or list of matrices) is present in the `R` environment rather than multiple copies of it for each subset.

### `$matrix`

This is the multidimensional space, stored in the `R` environment as a `list` object containing one or more `matrix` objects.
Each `matrix` requires row names but not column names (optional).
By default, if the row names are missing, `dispRity` function will arbitrarily generate them in numeric order (i.e. `rownames(matrix) <- 1:nrow(matrix)`).
This element of the `dispRity` object is never modified.

### `$call`

This element contains the information on the `dispRity` object content.
It is a `list` that can contain the following:

 * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`), their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`) and eventual information about the trees and matrices used through `chrono.subsets`. This element generated only once via `chrono.subsets()` and `custom.subsets()`.
 * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use. This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
 * `$call$bootstrap`: this is a `list` containing three elements:
    * `[[1]]`: the number of bootstrap replicates (`numeric`)
    * `[[2]]`: the bootstrap method (`character`)
    * `[[3]]`: the rarefaction levels (`numeric` vector)
 * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`. These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).

### `$subsets`

This element contain the eventual subsets of the multidimensional space.
It is a `list` of subset names.
Each subset name is in turn a `list` of at least one element called `elements` which is in turn a `matrix`.
This `elements` matrix is the raw (observed) elements in the subsets.
The `elements` matrix is composed of `numeric` values in one column and _n_ rows (the number of elements in the subset).
Each of these values are a "pointer" (`C` inspired) to the element of the `$matrix`.
For example, lets assume a `dispRity` object called `disparity`, composed of at least one subsets called `sub1`:

@@

When a probabilistic model was used to select the elements (models that have the `"split"` suffix, e.g. `chrono.subsets(..., model = "gradual.split")`), the `$elements` is a matrix containing a pair of elements of the matrix and a probability for sampling the first element in that list:

```
 disparity$subsets$sub1$elements
      [,1] [,2]       [,3]
 [1,]   73   36 0.01871893
 [2,]   74   37 0.02555876
 [3,]   33   38 0.85679821
```

In this example, you can read the table row by row as: "there is a probability of `0.018` for sampling element `73` and a probability of `0.82` (`1-0.018`) of sampling element `36`".

### `$disparity`

The `$disparity` element is identical to the `$subsets` element structure (a list of list(s) containing matrices) but the matrices don't contain "pointers" to `$matrix` but the disparity result of the disparity metric applied to the "pointers".
For example, in our first example (`$elements`) from above, if the disparity metric is of dimensions level 1, we would have:

@@
